---
title: "Regression Analysis - MGT6203 Project"
author: "bella"
date: "2024-10-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r}
# import libraries
library(ggplot2) 
library(tidyverse) 
library(broom)
library(dplyr)
library(tidyverse)
library(ggpmisc)
library(zoo)
library(car)
library(MASS)
library(kableExtra)


library(lars)
library(MASS)
library(glmnet)
library(grpreg)
library(scales)
library(caret)

options(dplyr.summarise.inform = FALSE) 
```

### Load and Pre-process data
```{r}
# Set seed for reproducibility
set.seed(123)

# read data
ep.data <- read.csv("employee_performance.csv", header=TRUE)
ep.data$Job_Satisfaction = as.factor(ep.data$Job_Satisfaction)
ep.data$Work_Life_Balance = as.factor(ep.data$Work_Life_Balance)
ep.data$Promotions_Last_5_Years = as.factor(ep.data$Promotions_Last_5_Years)

# train/test split
trainRows = sample.int(nrow(ep.data),size=ceiling(0.8*nrow(ep.data)))

trainData = ep.data[trainRows,]
testData = ep.data[-trainRows,]
```


### Full Regression

Fit full regression using training set
```{r}
set.seed(123)
model.full <- lm(Performance_Score~., data=trainData)
summary(model.full)
```

#### Overall Significance

The F-test is used to test for overall model significance with alpha=.01:

  - null hypothesis: all the regression coefficients are zero
  - alternative hypothesis: at least on regression coefficient is not zero

Since the p-value for the F-test is less than 0.01, we can reject the null hypothesis and conclude the model is significant at the 0.01 significance level.

### Goodness of Fit Assessment
```{r}
# add standardized residuals and fitted values 
fittedData <- trainData
fittedData$Standardized_Residuals <- rstandard(model.full)
fittedData$Fitted_Values <- model.full$fitted.values
```

**1. LINEARITY**
```{r}
# plot standardized residuals model.full vs. the quantitative predictors to assess if linearity assumption holds for the predictors
```

**2. CONSTANT VARIANCE and 3. INDEPENDENCE (UNCORRELATED ERRORS)**
```{r}
# Fitted values Scatter Plot
ggplot(data=fittedData, aes(x=Fitted_Values, y=Standardized_Residuals)) +
  geom_point(color='black') +
  ggtitle("Standardized Residuals vs Fitted Values") +
  geom_hline(yintercept=0, color="red", size=0.3)
```

**4. NORMALITY**
```{r}
# histogram of standardized residuals: check normality, constant var, uncorr errors
std.resids <- rstandard(model.full)
hist(std.resids,main="Histogram of Standardized Residuals",  xlab="Standardized Residuals")

# Q-Q plot: check normality
qqPlot(std.resids, main="Normal QQ-Plot Standardized of Residuals", envelope=0.95,
      xlab="Theoretical Quantiles", ylab="Sample Quantiles")
```

### Multi-collinearity

```{r}
vif.threshold <- 5

vif.model.full <- round(vif(model.full),3)
vif.model.full <- as.data.frame(vif.model.full)
vif.model.full$exceedsThreshold <- ifelse(vif.model.full$GVIF > vif.threshold, "True", "False")
kable(vif.model.full, caption = "Model1 VIFs")
```

### Outliers
```{r}

cooks.threshold <- 1

# plot Cook’s distances for model1
cook = cooks.distance(model.full)
plot(cook, type="h", lwd=3,col="darkorange",
  ylab = "Cook's Distance", main="Cook's Distance Plot")
abline(h = cooks.threshold, lty = 2, col = "darkblue")
```

### Stepwise Variable Selection: Forward, Backward, Forward-Backward

minimum model: The minimum model used for stepwise regression was the intercept-only model. This is because there are not controlling variables that we must enter in the model along with the intercept. 

selection criteria: The AIC was used as the criteria to add variables in forward stepwise and remove variables in backward regression. 

Forward-selection:

  - AIC: 59485.7 
  - BIC: 59555.57
  - variables selected: (Intercept), Experience, Work_Hours_Per_Week, Age, Education_LevelBachelor, Education_LevelHigh School, Education_LevelMaster, Education_LevelPhD, Annual_Bonus

Backward-selection:

  - AIC: AIC: 59485.7 
  - BIC: 59555.57
  - variables selected: (Intercept), Experience, Work_Hours_Per_Week, Age, Education_LevelBachelor, Education_LevelHigh School, Education_LevelMaster, Education_LevelPhD, Annual_Bonus
  
Forward-Backward Model:
  - AIC: 59485.7 
  - BIC: 59555.57
  - variables selected: (Intercept), Experience, Work_Hours_Per_Week, Age, Education_LevelBachelor, Education_LevelHigh School, Education_LevelMaster, Education_LevelPhD, Annual_Bonus

#### Forward-Selection

```{r}
### Apply Stepwise to training data
full.model <- lm(Performance_Score~., data=trainData)

minimum.model <- lm(Performance_Score~1, data=trainData)

# Forward
stepwise_forward <- step(minimum.model ,scope = list(lower=minimum.model,upper=full.model), direction = "forward", k=2)

summary(stepwise_forward)
```

```{r}
n = nrow(trainData)
# selected variables
coef(stepwise_forward)

# AIC and BIC
cat("AIC:", AIC(stepwise_forward, k=2),"\n")
cat("BIC:", AIC(stepwise_forward, k=log(n)))
```

### Backward-Selection
```{r}
# Backward
stepwise_backward <- step(minimum.model ,scope = list(lower=minimum.model,upper=full.model), direction = "forward", k=2)
```

```{r}
# selected variables
coef(stepwise_backward)

# AIC and BIC
cat("AIC:", AIC(stepwise_backward, k=2),"\n")
cat("BIC:", AIC(stepwise_backward, k=log(n)))
```

#### Both (Forward and Backward)
```{r}
# Both
fwd_bkwd_model<- step(minimum.model ,scope = list(lower=minimum.model,upper=full.model), direction = "forward", k=2)
```

```{r}
# filter for coefficients with a p-value <0.01
which(summary(fwd_bkwd_model)$coeff[,4]<0.01)
```

```{r}
# selected variables
coef(fwd_bkwd_model)

# AIC and BIC
cat("AIC:", AIC(fwd_bkwd_model, k=2),"\n")
cat("BIC:", AIC(fwd_bkwd_model, k=log(n)))
```
### Model Comparison

```{r}
stepwise_model = lm(Performance_Score ~ Experience+Work_Hours_Per_Week+Age+Education_Level+Annual_Bonus, data=trainData)

stepwise_signif_only = lm(Performance_Score ~ Experience+Work_Hours_Per_Week+Age+Education_Level, data=trainData)
```

```{r}
# compare full and stepwise
anova(model.full, stepwise_model)
```
For anova(model.full, stepwise): The p-value is large in the ANOVA test comparing the full regression to the model selected with stepwise. This means we cannot reject the null hypothesis. The two models are not statistically significantly different at the 0.05 level. Hence, choose the reduced model becuase it has fewer predicting variables.

```{r}
anova(stepwise_model, stepwise_signif_only)
```

For anova(stepwise, signif_stepwise): The p-value is less than 0.05. This means there's at least one variable in stepwise_model that will improve the explanatory power of the model. Hence, choose stepwise_model over further reduced one.

### Regularized Regression for Variable Selection:

LASSO. Group LASSO, and Elastic Net are three regularized regression models that can be used for regularized regression. Elastic Net often outperforms LASSO, since Elastic Net leverages the L1 and L2 penalties, so it has the advantages of both LASSO and Ridge regression, whereas LASSO only uses the L1 penalty. 

glmnet() will re-scale the predictor variables for regularized regression
```{r}
set.seed(123)
# format predictors and response 
x.train <- model.matrix(Performance_Score ~ ., trainData)[,-1]
y.train <- trainData$Performance_Score
```

### Lasso
```{r}
# LASSO
set.seed(123)
## Find optimal lambda with 10-fold CV 
lasso_model.cv = cv.glmnet(x.train,  y.train, alpha=1, nfolds=10)

cat("optimal lambda:", lasso_model.cv$lambda.min)
```

```{r}
set.seed(123)
## Fit lasso model with 100 values for lambda
lasso_model = glmnet(x.train, y.train, alpha=1, nlambda=100)

## coefficients at optimal lambda
coef(lasso_model, s=lasso_model.cv$lambda.min)

## Plot coefficient paths
plot(lasso_model , xvar="lambda", lwd=2, label=TRUE)
abline(v=log(lasso_model.cv$lambda.min), col='black', lty=2, lwd=2)
```

### Elastic Net
```{r}
set.seed(123)
# Find the optimal lambda using 10-fold CV for ELASTIC NET
elastic_net_model.cv <- cv.glmnet(x.train, y.train, alpha=0.5, nfolds=10)

cat("optimal lambda:", elastic_net_model.cv$lambda.min)
 
# fit model for 100 values for lambda
elastic_net_model <- glmnet(x = x.train, y = y.train, alpha=0.5, nlambda=100)

## Extract coefficients at optimal lambda
coef(elastic_net_model, s=elastic_net_model.cv$lambda.min)
```

```{r}
set.seed(123)
## Plot coefficient paths
plot(elastic_net_model, xvar="lambda", lwd=2)
abline(v=log(elastic_net_model.cv$lambda.min), col='black', lty=2, lwd=2)
```

### Decision Tree

```{r}
library(rpart)
library(rpart.plot)
set.seed(123)
# Build the decision tree model
decision_tree_model <- rpart(Performance_Score ~ ., data = trainData, method = "anova")

# Print the model summary
summary(decision_tree_model)
```

```{r}
rpart.plot(decision_tree_model, type = 3, fallen.leaves = TRUE, cex = 0.7)
```

### Random Forest
```{r}
set.seed(123)
library(randomForest)

# for random forest
rf_model <- randomForest(trainData$Performance_Score~.,data = trainData)

```

### Make Predictions

```{r}
## PREDICTIONS
y.test <- testData$Performance_Score

# full model
predict.full <- predict(model.full,newdata=testData)

# stepwise
predict.stepwise <- predict(stepwise_model,newdata=testData)

# LASSO
# Retrain OLS using Lasso-selected predictors for better coeff estimates
index.lasso <- which(coef(lasso_model, lasso_model.cv$lambda.min) == 0)
lasso.predictors <- as.data.frame(x.train)[-(index.lasso-1)]
lasso.retrained <- glm(y.train ~ ., data = lasso.predictors)
# Set test data to correct format
new_test <- model.matrix( ~ ., data=testData)[,-1]
# get predicted probabilities for the test set
predict.lasso = predict(lasso.retrained, newdata = as.data.frame(new_test),
                     type = "response")

# decision tree
predict.dtree <- predict(decision_tree_model, testData)

# random forest
predict.rf <- predict(rf_model, testData, type="response")

# function for predictions
pred_metrics = function(model_name, actual, pred){
  cat("\n", model_name, "\n")
  mspe = mean((actual - pred)^2)
  mae = mean(abs(actual-pred))
  cat("mspe:", mspe, "\n", "mae:", mae, "\n")
}

# metrics for full model
pred_metrics("Full Regression", y.test, predict.full)

# metrics for forward stepwise
pred_metrics("Stepwise", y.test, predict.stepwise)

# metrics for lasso
pred_metrics("LASSO", y.test, predict.lasso)

# metrics for decision tree
pred_metrics("Decision Tree", y.test, predict.dtree)

# metrics for random forest
pred_metrics("Random Forest Model", y.test, predict.rf)


# adjusted r-squared for training set
# summary(model.full)$adj.r.squared
# summary(stepwise_model)$adj.r.squared

```

