---
title: "Regression Analysis - MGT6203 Project"
author: "bella"
date: "2024-10-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r}
# import libraries
library(ggplot2) 
library(tidyverse) 
library(broom)
library(dplyr)
library(tidyverse)
library(ggpmisc)
library(zoo)
library(car)
library(MASS)
library(kableExtra)


library(lars)
library(MASS)
library(glmnet)
library(grpreg)
library(scales)
library(caret)

options(dplyr.summarise.inform = FALSE) 
```

### Load and Pre-process data
```{r}
# Set seed for reproducibility
set.seed(123)

# read data
ep.data <- read.csv("employee_performance.csv", header=TRUE)
ep.data$Job_Satisfaction = as.factor(ep.data$Job_Satisfaction)
ep.data$Work_Life_Balance = as.factor(ep.data$Work_Life_Balance)
ep.data$Promotions_Last_5_Years = as.factor(ep.data$Promotions_Last_5_Years)

# train/test split
trainRows = sample.int(nrow(ep.data),size=ceiling(0.8*nrow(ep.data)))

trainData = ep.data[trainRows,]
testData = ep.data[-trainRows,]
```


### Full Regression

Fit full regression using training set
```{r}

model.full <- lm(Performance_Score~., data=trainData)
summary(model.full)
```

#### Overall Significance

The F-test is used to test for overall model significance with alpha=.01:

  - null hypothesis: all the regression coefficients are zero
  - alternative hypothesis: at least on regression coefficient is not zero

Since the p-value for the F-test is less than 0.01, we can reject the null hypothesis and conclude the model is significant at the 0.01 significance level.

### Goodness of Fit Assessment
```{r}
# add standardized residuals and fitted values 
fittedData <- trainData
fittedData$Standardized_Residuals <- rstandard(model.full)
fittedData$Fitted_Values <- model.full$fitted.values
```

**1. LINEARITY**
```{r}
# plot standardized residuals model.full vs. the quantitative predictors to assess if linearity assumption holds for the predictors
```

**2. CONSTANT VARIANCE and 3. INDEPENDENCE (UNCORRELATED ERRORS)**
```{r}
# Fitted values Scatter Plot
ggplot(data=fittedData, aes(x=Fitted_Values, y=Standardized_Residuals)) +
  geom_point(color='black') +
  ggtitle("Standardized Residuals vs Fitted Values") +
  geom_hline(yintercept=0, color="red", size=0.3)
```

**4. NORMALITY**
```{r}
# histogram of standardized residuals: check normality, constant var, uncorr errors
std.resids <- rstandard(model.full)
hist(std.resids,main="Histogram of Standardized Residuals",  xlab="Standardized Residuals")

# Q-Q plot: check normality
qqPlot(std.resids, main="Normal QQ-Plot Standardized of Residuals", envelope=0.95,
      xlab="Theoretical Quantiles", ylab="Sample Quantiles")
```

### Multi-collinearity

```{r}
vif.threshold <- 5

vif.model.full <- round(vif(model.full),3)
vif.model.full <- as.data.frame(vif.model.full)
vif.model.full$exceedsThreshold <- ifelse(vif.model.full$GVIF > vif.threshold, "True", "False")
kable(vif.model.full, caption = "Model1 VIFs")
```

### Outliers
```{r}

cooks.threshold <- 1

# plot Cook’s distances for model1
cook = cooks.distance(model.full)
plot(cook, type="h", lwd=3,col="darkorange",
  ylab = "Cook's Distance", main="Cook's Distance Plot")
abline(h = cooks.threshold, lty = 2, col = "darkblue")
```

### Stepwise Variable Selection: Forward, Backward, Forward-Backward

minimum model: The minimum model used for stepwise regression was the intercept-only model. This is because there are not controlling variables that we must enter in the model along with the intercept. 

selection criteria: The AIC was used as the criteria to add variables in forward stepwise and remove variables in backward regression. 

Forward-selection:

  - AIC: 59485.7 
  - BIC: 59555.57
  - variables selected: (Intercept), Experience, Work_Hours_Per_Week, Age, Education_LevelBachelor, Education_LevelHigh School, Education_LevelMaster, Education_LevelPhD, Annual_Bonus

Backward-selection:

  - AIC: AIC: 59485.7 
  - BIC: 59555.57
  - variables selected: (Intercept), Experience, Work_Hours_Per_Week, Age, Education_LevelBachelor, Education_LevelHigh School, Education_LevelMaster, Education_LevelPhD, Annual_Bonus
  
Forward-Backward Model:
  - AIC: 59485.7 
  - BIC: 59555.57
  - variables selected: (Intercept), Experience, Work_Hours_Per_Week, Age, Education_LevelBachelor, Education_LevelHigh School, Education_LevelMaster, Education_LevelPhD, Annual_Bonus

#### Forward-Selection

```{r}
### Apply Stepwise to training data
full.model <- lm(Performance_Score~., data=trainData)

minimum.model <- lm(Performance_Score~1, data=trainData)

# Forward
stepwise_forward <- step(minimum.model ,scope = list(lower=minimum.model,upper=full.model), direction = "forward", k=2)
```

```{r}
# selected variables
coef(stepwise_forward)

# AIC and BIC
cat("AIC:", AIC(stepwise_forward),"\n")
cat("BIC:", BIC(stepwise_forward))
```

### Backward-Selection
```{r}
# Backward
stepwise_backward <- step(minimum.model ,scope = list(lower=minimum.model,upper=full.model), direction = "forward", k=2)
```

```{r}
# selected variables
coef(stepwise_backward)

# AIC and BIC
cat("AIC:", AIC(stepwise_backward),"\n")
cat("BIC:", BIC(stepwise_backward))
```

#### Both (Forward and Backward)
```{r}
# Both
fwd_bkwd_model<- step(minimum.model ,scope = list(lower=minimum.model,upper=full.model), direction = "forward", k=2)
```

```{r}
# selected variables
coef(fwd_bkwd_model)

# AIC and BIC
cat("AIC:", AIC(fwd_bkwd_model),"\n")
cat("BIC:", BIC(fwd_bkwd_model))
```
### Regularized Regression for Variable Selection:

LASSO. Group LASSO, and Elastic Net are three regularized regression models that can be used for regularized regression. Elastic Net often outperforms LASSO, since Elastic Net leverages the L1 and L2 penalties, so it has the advantages of both LASSO and Ridge regression, whereas LASSO only uses the L1 penalty. 

The data must be scaled before applying regularized regression, so we used scale() to rescale the training data.

### Lasso
```{r}
## Scale predictors and response
num.variables <- cbind(Age=trainData$Age, Experience=trainData$Experience, Salary=trainData$Salary,
           Job_Satisfaction=trainData$Job_Satisfaction, Work_Life_Balance=trainData$Work_Life_Balance,
           Annual_Bonus=trainData$Annual_Bonus, Commute_Time=trainData$Commute_Time,
           Work_Hours_Per_Week=trainData$Work_Hours_Per_Week, 
           Promotions_Last_5_Years=trainData$Promotions_Last_5_Years,
           Training_Hours_Last_Year=trainData$Training_Hours_Last_Year,
           Overtime_Hours_Per_Week=trainData$Overtime_Hours_Per_Week)

#  ep.data$Education_Level, ep.data$Department

scaled.num.variables <- scale(num.variables)
#response.scaled <- scale(trainData$Performance_Score)

num_var_scale <- scale(num.variables) 
dv <- dummyVars("~ Education_Level+ Department", data = trainData)
# 
#num.scaled.matrix <- matrix(scaled.num.variables, nrow = nrow(num.variables), byrow = FALSE)


# Create the dummy variables dataframe
x_dummy <- predict(dv, newdata = trainData)
# 
x_pred_scale <- cbind(x_dummy, as.matrix(num_var_scale))
```

```{r}
object = lars(x=x_pred_scale, y= trainData$Performance_Score)
object

round(object$Cp, 2)
plot(object)

plot.lars(object, xvar="df", plottype="Cp")
```

```{r}
# LASSO
set.seed(123)
## Find optimal lambda with 10-fold CV 
empPerf.model.cv = cv.glmnet(x_pred_scale,  trainData$Performance_Score, alpha=1, nfolds=10)

cat("optimal lambda:", empPerf.model.cv$lambda.min)
```

```{r}
## Fit lasso model with 100 values for lambda
empPerf.model = glmnet(x_pred_scale, trainData$Performance_Score, alpha=1, nlambda=100)

## coefficients at optimal lambda
coef(empPerf.model, s=empPerf.model.cv$lambda.min)

## Plot coefficient paths
plot(empPerf.model , xvar="lambda", lwd=2)
abline(v=log(empPerf.model.cv$lambda.min), col='black', lty=2, lwd=2)
```
### Decision Tree

```{r}
library(rpart)
library(rpart.plot)

# Build the decision tree model
decision_tree_model <- rpart(Performance_Score ~ ., data = trainData, method = "anova")

# Print the model summary
summary(decision_tree_model)
```

```{r}
rpart.plot(decision_tree_model, type = 3, fallen.leaves = TRUE, cex = 0.7)
```

# Decision tree using the tree package
```{r}
# Loads in the library
library(tree)

# Builds the decision tree model
decision_tree_ep <- tree(Performance_Score~., data = trainData)

```

```{r}
# Prints the model summary
summary(decision_tree_ep)
```

```{r}
# Plots the decision tree
plot(decision_tree_ep)
text(decision_tree_ep,  cex=0.6)
```





# Random Forest
```{r}
# Loads the necesary library
library(randomForest)

# Sets the seed for reproducibility
set.seed(123)

# Builds the random forest model
random_forest_model <- randomForest(Performance_Score~., data=trainData, ntree=100)

# Prints the model summary
print(random_forest_model)
```


```{r}
# Views the importance of each predictor
importance(random_forest_model)

```


```{r}
# Plots the importance of predictors
varImpPlot(random_forest_model, cex = 0.8)

```






