---
title: "MGT6203_group_project_outline"
author: "bella"
date: "2024-10-09"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message = FALSE, warning = FALSE}
# Import libraries
library(ggplot2) 
library(tidyverse) 
library(broom)
library(dplyr)
library(tidyverse)
library(ggpmisc)
library(zoo)
library(car)
library(reshape2)
library(corrplot)
options(dplyr.summarise.inform = FALSE) 
```


```{r}
# Reads in the data
ep.data <- read.csv("employee_performance.csv", 
                    header=TRUE)
```

### Data Collection
  - any missing data?
  - if missing data, will we interpolate or remove observation?

Checks for missing values
```{r}
sum(is.na(ep.data))
```


Checks the structure of the data
```{r}
str(ep.data)
```

Provides the summary statistics of the data
```{r}
summary(ep.data)
```


Prints out the first few rows of the data
```{r}
head(ep.data)
```

### Exploratory Data Analysis (EDA)


- correlation matrix


Prepares the data
```{r}
ep.data_numeric <- ep.data[c("Age","Salary", "Experience","Job_Satisfaction", 
                             "Work_Life_Balance", "Annual_Bonus","Commute_Time",
                             "Work_Hours_Per_Week", "Promotions_Last_5_Years",
                             "Training_Hours_Last_Year", "Overtime_Hours_Per_Week",
                             "Performance_Score")]
head(ep.data_numeric)
```


Computes the correlation matrix
```{r}
ep.cormatrix <- round(cor(ep.data_numeric), 2)

# Print the correlation matrix
print(ep.cormatrix)
```

Creates the heatmap with the corresponding correlation coefficients
```{r}
corrplot(ep.cormatrix, 
         method = "color",               # Use color for the heatmap
         col = colorRampPalette(c("purple4", "white",
                                  "violet"))(200),  # Color gradient
         type = "full",                  # Show both upper and lower triangles
         order = "original",               # Order by hierarchical clustering
         addCoef.col = "black",          # Add correlation coefficients in black
         tl.col = "black",               # Text label color
         tl.srt = 90,                    # Text label rotation
         diag = TRUE,                    # Show the diagonal
         number.cex = 0.55, # Size of the correlation coefficient numbers
         tl.cex = 0.55)
```


Creates a correlation heatmap with numbers using corrplot
```{r}
corrplot(ep.cormatrix, 
         col = colorRampPalette(c("lightblue", "violet", 
                                  "darkblue"))(200),
         number.cex = 0.4, 
         tl.cex = 0.6)
```



1. Overall Correlation for the Numeric values of the Employee dataset. 

```{r}
# Select numeric columns for the correlation matrix
ep.numeric_vars <- ep.data %>%
  dplyr::select(Age, Salary, Experience, 
                Job_Satisfaction, 
                Work_Life_Balance, 
         Annual_Bonus, Commute_Time, 
         Work_Hours_Per_Week, 
         Promotions_Last_5_Years, 
         Training_Hours_Last_Year, Overtime_Hours_Per_Week,
         Performance_Score)

# Computes the correlation matrix
ep.overall_cormatrix <- round(cor(ep.numeric_vars), 2)

# Prints out the correlation matrix
print(ep.overall_cormatrix)
```


Creates the heatmap with the overall corresponding correlation coefficients
```{r}
corrplot(ep.overall_cormatrix, 
         method = "color",               # Use color for the heatmap
         col = colorRampPalette(c("darkblue", "white", 
                                  "lightblue3"))(200),  # Color gradient
         type = "full",                  # Show both upper and lower triangles
         order = "original",               # Order by hierarchical clustering
         addCoef.col = "black",          # Add correlation coefficients in black
         tl.col = "black",               # Text label color
         tl.srt = 90,                    # Text label rotation
         diag = TRUE,                    # Show the diagonal
         number.cex = 0.55, # Size of the correlation coefficient numbers
         tl.cex = 0.6)
```



Creates a correlation heatmap with numbers using corrplot

```{r}
corrplot(ep.overall_cormatrix, 
         col = colorRampPalette(c("blue", 
                                  "lightgreen", "brown"))(200),
         number.cex = 0.4, 
         tl.cex = 0.6)

```




2. Correlation Matrices for Different Groups (Based on Categorical Variables)


2.1 Correlation matrix by Department


```{r}
# Defines the numeric columns
ep.numeric_columns <- c("Age", "Salary", "Experience", 
                        "Job_Satisfaction", "Work_Life_Balance", 
                     "Annual_Bonus", "Commute_Time", 
                     "Work_Hours_Per_Week", 
                     "Promotions_Last_5_Years", 
                     "Training_Hours_Last_Year", 
                     "Overtime_Hours_Per_Week", 
                     "Performance_Score")

# Split data by Department 
Department <- unique(ep.data$Department)

# Loop through each department level and calculate the correlation 
# matrix for numeric variables
for(departments in Department) {
  # Filter the data for each department 
  ep.department_data <- ep.data %>% filter(Department == departments)
  
  # Subset the data to only include numeric columns
  ep.department_data_numeric <- ep.department_data %>% 
    dplyr::select(all_of(ep.numeric_columns))
  
  # Calculate the correlation matrix
  ep.department_cormatrix <- round(cor(ep.department_data_numeric), 2)
  

}

```



```{r}
# Defines the variables needed for the analysis
ep.engineering_data <- ep.data %>% filter(Department == "Engineering") %>% 
  dplyr::select(all_of(ep.numeric_columns))
ep.engineering_cormatrix <- round(cor(ep.engineering_data), 2)
print(ep.engineering_cormatrix)
```


Creates a heatmap with the corresponding correlation coefficients for the Engineering Department
```{r}
corrplot(ep.engineering_cormatrix, 
         method = "color",               # Use color for the heatmap
         col = colorRampPalette(c("yellow", 
                                  "white", 
                                  "darkorange"))(200),  # Color gradient
         type = "full",                  # Show both upper and lower triangles
         order = "original",               # Order by hierarchical clustering
         addCoef.col = "black",          # Add correlation coefficients in black
         tl.col = "black",               # Text label color
         tl.srt = 90,                    # Text label rotation
         diag = TRUE,                    # Show the diagonal
         number.cex = 0.55, # Size of the correlation coefficient numbers
         tl.cex = 0.5)
```

Create a correlation heatmap with numbers using corrplot
```{r}
corrplot(ep.engineering_cormatrix, 
         col = colorRampPalette(c("yellow", 
                                  "green", "blue"))(200),
         number.cex = 0.4, 
         tl.cex = 0.6)
```

Defines the variables needed to create the Correlation Matrix and Heatmap for the Finance Department
```{r}
# Finance Department Correlation Matrix and Heatmap
ep.finance_data <- ep.data %>% filter(Department == "Finance") %>% 
  dplyr::select(all_of(ep.numeric_columns))
ep.finance_cormatrix <- round(cor(ep.finance_data), 2)
print(ep.finance_cormatrix)
```

Creates a heatmap with the corresponding correlation coefficients for the Finance Department
```{r}
corrplot(ep.finance_cormatrix, 
         method = "color",               # Use color for the heatmap
         col = colorRampPalette(c("lightblue", 
                                  "white", "cyan"))(200),  # Color gradient
         type = "full",                  # Show both upper and lower triangles
         order = "original",               # Order by hierarchical clustering
         addCoef.col = "black",          # Add correlation coefficients in black
         tl.col = "black",               # Text label color
         tl.srt = 90,                    # Text label rotation
         diag = TRUE,                    # Show the diagonal
         number.cex = 0.55, # Size of the correlation coefficient numbers
         tl.cex = 0.6)
```

Create a correlation heatmap with numbers using corrplot
```{r}
corrplot(ep.finance_cormatrix, 
         col = colorRampPalette(c("yellow", 
                                  "gold", "darkorange"))(200),
         number.cex = 0.4, 
         tl.cex = 0.6)
```


Defines the variables needed to create the Correlation Matrix and Heatmap for the HR Department
```{r}
# HR Department
ep.hr_data <- ep.data %>% filter(Department == "HR") %>% 
  dplyr::select(all_of(ep.numeric_columns))
ep.hr_cormatrix <- round(cor(ep.hr_data), 2)
print(ep.hr_cormatrix)
```

Creates a heatmap with the corresponding correlation coefficients for the HR Department
```{r}
corrplot(ep.hr_cormatrix, 
         method = "color",               # Use color for the heatmap
         col = colorRampPalette(c("violet", "white",
                                  "red1"))(200),  # Color gradient
         type = "full",                  # Show both upper and lower triangles
         order = "original",               # Order by hierarchical clustering
         addCoef.col = "black",          # Add correlation coefficients in black
         tl.col = "black",               # Text label color
         tl.srt = 90,                    # Text label rotation
         diag = TRUE,                    # Show the diagonal
         number.cex = 0.55, # Size of the correlation coefficient numbers
         tl.cex = 0.6)
```

Create a correlation heatmap with numbers using corrplot
```{r}
corrplot(ep.hr_cormatrix, 
         col = colorRampPalette(c("orange", "green", 
                                  "purple"))(200),
         number.cex = 0.4, 
         tl.cex = 0.55)
```



Defines the variables needed to create the Correlation Matrix and Heatmap for the Marketing Department
```{r}
# Marketing Department
ep.marketing_data <- ep.data %>% filter(Department == "Marketing") %>% 
  dplyr::select(all_of(ep.numeric_columns))
ep.marketing_cormatrix <- round(cor(ep.marketing_data), 2)
print(ep.marketing_cormatrix)
```



Creates a heatmap with the corresponding correlation coefficients for the Marketing Department
```{r}
corrplot(ep.marketing_cormatrix, 
         method = "color",               # Use color for the heatmap
         col = colorRampPalette(c("yellow", "white", 
                                  "brown"))(200),  # Color gradient
         type = "full",                  # Show both upper and lower triangles
         order = "original",               # Order by hierarchical clustering
         addCoef.col = "black",          # Add correlation coefficients in black
         tl.col = "black",               # Text label color
         tl.srt = 90,                    # Text label rotation
         diag = TRUE,                    # Show the diagonal
         number.cex = 0.55, # Size of the correlation coefficient numbers
         tl.cex = 0.6)
```



Create a correlation heatmap with numbers using corrplot
```{r}
corrplot(ep.marketing_cormatrix, 
         col = colorRampPalette(c("yellow", "darkorange",
                                  "darkgreen"))(200),
         tl.cex = 0.6)
```




Defines the variables needed to create the Correlation Matrix and Heatmap for the Operations Department
```{r}
# Operations Department
ep.operations_data <- ep.data %>% filter(Department == "Operations") %>% 
  dplyr::select(all_of(ep.numeric_columns))
ep.operations_cormatrix <- round(cor(ep.operations_data), 2)
print(ep.operations_cormatrix)
```


Creates a heatmap with the corresponding correlation coefficients for the Operations Department
```{r}
corrplot(ep.operations_cormatrix, 
         method = "color",               # Use color for the heatmap
         col = colorRampPalette(c("cyan", 
                                  "white", "cyan4"))(200),  # Color gradient
         type = "full",                  # Show both upper and lower triangles
         order = "original",               # Order by hierarchical clustering
         addCoef.col = "black",          # Add correlation coefficients in black
         tl.col = "black",               # Text label color
         tl.srt = 90,                    # Text label rotation
         diag = TRUE,                    # Show the diagonal
         number.cex = 0.55, # Size of the correlation coefficient numbers
         tl.cex = 0.6)
```


Create a correlation heatmap with numbers using corrplot
```{r}
corrplot(ep.operations_cormatrix, 
         col = colorRampPalette(c("yellow", 
                                  "lightgreen", "blue4"))(200),
         tl.cex = 0.6)
```




Defines the variables needed to create the Correlation Matrix and Heatmap for the Sales Department
```{r}
# Sales Department
ep.sales_data <- ep.data %>% filter(Department == "Sales") %>% 
  dplyr::select(all_of(ep.numeric_columns))
ep.sales_cormatrix <- round(cor(ep.sales_data), 2)
print(ep.sales_cormatrix)
```

Creates a heatmap with the corresponding correlation coefficients for the Sales Department
```{r}
corrplot(ep.sales_cormatrix, 
         method = "color",               # Use color for the heatmap
         col = colorRampPalette(c("deepskyblue", 
                                  "white", "dodgerblue"))(200),  # Color gradient
         type = "full",                  # Show both upper and lower triangles
         order = "original",               # Order by hierarchical clustering
         addCoef.col = "black",          # Add correlation coefficients in black
         tl.col = "black",               # Text label color
         tl.srt = 90,                    # Text label rotation
         diag = TRUE,                    # Show the diagonal
         number.cex = 0.55, # Size of the correlation coefficient numbers
         tl.cex = 0.6)
```

Create a correlation heatmap with numbers using corrplot
```{r}
corrplot(ep.sales_cormatrix, 
         col = colorRampPalette(c("violet", 
                                  "deepskyblue", "blue"))(200),
         tl.cex = 0.6)
```


2.2 Correlation Matrix by Education level
```{r}
# Define numeric columns 
ep.numeric_columns <- c("Age", "Salary", "Experience", "Job_Satisfaction", 
                        "Work_Life_Balance", 
                     "Annual_Bonus", "Commute_Time", 
                     "Work_Hours_Per_Week", 
                     "Promotions_Last_5_Years", 
                     "Training_Hours_Last_Year", 
                     "Overtime_Hours_Per_Week", 
                     "Performance_Score")

# Split data by Education Level
ep.education_levels <- unique(ep.data$Education_Level)

# Loop through each education level and calculate the correlation matrix for numeric variables
for(education in ep.education_levels) {
  # Filter the data for each education level
  ep.education_data <- ep.data %>% filter(Education_Level == education)
  
  # Subset the data to only include numeric columns
  ep.education_data_numeric <- ep.education_data %>% dplyr::select(all_of(ep.numeric_columns))
  
  # Calculate the correlation matrix
  ep.education_cormatrix <- round(cor(ep.education_data_numeric), 2)
  
}
```



Defines the variables needed to create the Correlation Matrix and Heatmap for the Associate Degree Education Level
```{r}
# Associate Degree Correlation Matrix and Heatmap
ep.associate_data <- ep.data %>% filter(Education_Level == "Associate") %>% 
  dplyr::select(all_of(ep.numeric_columns))
ep.associate_cormatrix <- round(cor(ep.associate_data), 2)
print(ep.associate_cormatrix)
```

Creates a heatmap with the corresponding correlation coefficients for the Associate Degree Education Level
```{r}
corrplot(ep.associate_cormatrix, 
         method = "color",               # Use color for the heatmap
         col = colorRampPalette(c("gold", "white", "gold3"))(200),  # Color gradient
         type = "full",                  # Show both upper and lower triangles
         order = "original",               # Order by hierarchical clustering
         addCoef.col = "black",          # Add correlation coefficients in black
         tl.col = "black",               # Text label color
         tl.srt = 90,                    # Text label rotation
         diag = TRUE,                    # Show the diagonal
         number.cex = 0.55, # Size of the correlation coefficient numbers
         tl.cex = 0.6)
```

Create a correlation heatmap with numbers using corrplot
```{r}
corrplot(ep.associate_cormatrix, 
         col = colorRampPalette(c("yellow", 
                                  "lightgreen", "deepskyblue"))(200),
         tl.cex = 0.55)
```


Defines the variables needed to create the Correlation Matrix and Heatmap for the Bachelor's Degree Education Level
```{r}
# Bachelor Degree Correlation Matrix and Heatmap
ep.bachelor_data <- ep.data %>% filter(Education_Level == "Bachelor") %>% 
  dplyr::select(all_of(ep.numeric_columns))
ep.bachelor_cormatrix <- round(cor(ep.bachelor_data), 2)
print(ep.bachelor_cormatrix)
```

Creates a heatmap with the corresponding correlation coefficients for the Bachelor's Degree Education Level
```{r}
corrplot(ep.bachelor_cormatrix, 
         method = "color",               # Use color for the heatmap
         col = colorRampPalette(c("pink", 
                                  "white", "deeppink"))(200),  # Color gradient
         type = "full",                  # Show both upper and lower triangles
         order = "original",               # Order by hierarchical clustering
         addCoef.col = "black",          # Add correlation coefficients in black
         tl.col = "black",               # Text label color
         tl.srt = 90,                    # Text label rotation
         diag = TRUE,                    # Show the diagonal
         number.cex = 0.55, # Size of the correlation coefficient numbers
         tl.cex = 0.6)
```

Create a correlation heatmap with numbers using corrplot
```{r}
corrplot(ep.bachelor_cormatrix, 
         col = colorRampPalette(c("yellow", 
                                  "green", "darkgreen"))(200),
         tl.cex = 0.55)
```




Defines the variables needed to create the Correlation Matrix and Heatmap for the High School Degree Education Level
```{r}
# High School Correlation Matrix and Heatmap
ep.highschool_data <- ep.data %>% filter(Education_Level == "High School") %>% 
  dplyr::select(all_of(ep.numeric_columns))
ep.highschool_cormatrix <- round(cor(ep.highschool_data), 2)
print(ep.highschool_cormatrix)
```


Creates a heatmap with the corresponding correlation coefficients for the High School Degree Education Level
```{r}
corrplot(ep.highschool_cormatrix, 
         method = "color",               # Use color for the heatmap
         col = colorRampPalette(c("lightpink", 
                                  "white", 
                                  "mediumorchid"))(200),  # Color gradient
         type = "full",                  # Show both upper and lower triangles
         order = "original",               # Order by hierarchical clustering
         addCoef.col = "black",          # Add correlation coefficients in black
         tl.col = "black",               # Text label color
         tl.srt = 90,                    # Text label rotation
         diag = TRUE,                    # Show the diagonal
         number.cex = 0.55, # Size of the correlation coefficient numbers
         tl.cex = 0.6)
```


Create a correlation heatmap with numbers using corrplot
```{r}
corrplot(ep.highschool_cormatrix,
         col = colorRampPalette(c("yellow", 
                                  "orange", "darkred"))(200),
         tl.cex = 0.55)
```



Defines the variables needed to create the Correlation Matrix and Heatmap for the Master's Degree Education Level
```{r}
# Master Degree Correlation Matrix and Heatmap
ep.master_data <- ep.data %>% filter(Education_Level == "Master") %>% 
  dplyr::select(all_of(ep.numeric_columns))
ep.master_cormatrix <- round(cor(ep.master_data), 2)
print(ep.master_cormatrix)
```


Creates a heatmap with the corresponding correlation coefficients for the Master's Degree Education Level
```{r}
corrplot(ep.master_cormatrix, 
         method = "color",               # Use color for the heatmap
         col = colorRampPalette(c("salmon", 
                                  "white", 
                                  "salmon3"))(200),  # Color gradient
         type = "full",                  # Show both upper and lower triangles
         order = "original",               # Order by hierarchical clustering
         addCoef.col = "black",          # Add correlation coefficients in black
         tl.col = "black",               # Text label color
         tl.srt = 90,                    # Text label rotation
         diag = TRUE,                    # Show the diagonal
         number.cex = 0.55, # Size of the correlation coefficient numbers
         tl.cex = 0.6)
```


Create a correlation heatmap with numbers using corrplot
```{r}
corrplot(ep.master_cormatrix, 
         col = colorRampPalette(c("lightgreen", 
                                  "lightblue", "darkred"))(200),
         tl.cex = 0.55)
```


Defines the variables needed to create the Correlation Matrix and HEatmap for the PhD Degree Education Level
```{r}
# PhD Correlation Matrix and Heatmap
ep.phd_data <- ep.data %>% filter(Education_Level == "PhD") %>% 
  dplyr::select(all_of(ep.numeric_columns))
ep.phd_cormatrix <- round(cor(ep.phd_data), 2)
print(ep.phd_cormatrix)
```


Creates a heatmap with the corresponding correlation coefficients for the PhD Degree Education Level
```{r}
corrplot(ep.phd_cormatrix, 
         method = "color",               # Use color for the heatmap
         col = colorRampPalette(c("cadetblue1", 
                                  "white", 
                                  "cyan4"))(200),  # Color gradient
         type = "full",                  # Show both upper and lower triangles
         order = "original",               # Order by hierarchical clustering
         addCoef.col = "black",          # Add correlation coefficients in black
         tl.col = "black",               # Text label color
         tl.srt = 90,                    # Text label rotation
         diag = TRUE,                    # Show the diagonal
         number.cex = 0.55, # Size of the correlation coefficient numbers
         tl.cex = 0.6)
```


Create a correlation heatmap with numbers using corrplot
```{r}
corrplot(ep.phd_cormatrix, 
         col = colorRampPalette(c("yellow", "green",
                                  "blue"))(200),
         tl.cex = 0.55)
```














